A Fast and Tight Heuristic for A* in Road Networks

### 2021-01-18 paper/ch_potentials.tex
merge updates from Ben

\section{Introduction}
\label{sec:intro}
The past decade has seen a plethora of research on route planning in large street networks~\cite{bdgmpsww-rptn-16}.
Routing a user through a road network can be formalized as the shortest path problem in weighted graphs.
Nodes represent intersections.
Roads are modeled using edges.
Edges are weighted by their traversal times.
The problem can be solved with Dijkstra's algorithm~\cite{d-ntpcg-59}.
Unfortunately, on continental sized networks, it is too slow for many applications.
Thus, speed-up techniques have been developed.
One popular example are Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
They have been used successfully in many real world applications\todo{citation?}.
A CH exploits the inherent hierarchy of road networks.
In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant roads during query time.
A further popular example is Multi-Level-Dijkstra~(MLD)~\todo{pre CRP zitat?} also known as CRP~\cite{dgpw-crprn-13}.
It is also used in practice~\todo{citation, bing blog?}.
MLD also also uses shortcut edges.
Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.
% --- Unser Argument ist, dass die zu komplex zum erweitern sind... da passen diese 
%     Sätze nicht in die Story
% Their implementation complexity is also quite manageable.    
% This makes them very attractive in practice.

@@

Unfortunately, for many real world applications, this basic graph model is too simplistic.
Many additional features need to be considered.
% --- ich habe turn costs weg gelassen, weil wir die nicht in den experimenten haben, ich will keiner Reviewer auf blöde Ideeen bringen
% --- ich habe temporary driving bans weg gelassen, weil wir das nicht in den Experimenten haben und ich "we can do all of the above features" claimen möchte
This includes turn restrictions, live traffic, user preferences, traffic predictions.%, temporary driving bans for certain types of vehicles.
Specific applications often have additional application-specific requirements.
Extending Dijkstra's algorithm to support these features is usually easy.
On the other hand, while extending hierarchical speed-up techniques is possible, the algorithm development is vastly more complex.
For every feature, dedicated research paper(s) exist that extend CH.
Supporting the combination of several features is even harder. 
For example, we are not aware of any work combining all features mentioned above.
In this paper, we describe an algorithm that handles the combination of all mentioned features.

% --- mit "novel" und "propose" muss man vorsichtig sein, wegen SEA paper von letztem Jahr
We describe an algorithmic building block that mostly decouples feature development from the hierarchical speed-up algorithm.
Our goal is to simplify incorporating new features.
We build upon the A* algorithm \cite{A* literature}.
A* is a goal-directed variant of Dijkstra's algorithm.
It uses a heuristic\todo{heuristic vs potential} to guide the search towards the goal.
A heuristic is function that maps a node $v$ onto a lower bound of the distance from $v$ to the goal.
A*'s running time crucially depends on how tight this lower bound is.
Further, evaluating the heuristic must be fast.
The free flow travel time to the goal is a valid heuristic for most features.
In many settings, it is the tightest lower bound available during preprocessing, i.e., independent of real-time data and user settings.
However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
In this paper, we describe CH-Potentials, a technique that achieves both goals.
Internally, the heuristic uses a CH.
Fortunately, this is an implementation detail from the perspective of the A*.
To support a new feature, we only need to modify the A* algorithm.
The heuristic core containing the CH remains untouched.
Extending A* is vastly easier than extending a CH.
This enables us to design an algorithms that supports a multitude of features.
In addition, we describe optimizations for handling of low-degree nodes, common in road networks.
These low degree optimizations are applicable to Dijkstra's algorithm and A*.

@@

\subsection{Framework}

We consider a two-phase setup.
In the preprocessing phase, we are given a lower bound graph $G_\ell$ with lower bound weights $w_\ell$.
The lower bound weights are single scalar numbers for each edge.
We now perform some precomputation, i.e. apply the Contraction Hierarchy preprocessing to this graph.
This generates some auxiliary data which we will use in the query phase to efficiently compute distance estimates.
We will do so with the heuristic function $h_t(x)$ where $x$ and $t$ are nodes from $G_\ell$.
The heuristic function returns the shortest distance from $x$ to $t$ with respect to $w_\ell$.
Before each query, the heuristic has to be initialized with the target node $t$.
We will discuss the implementation of this heuristic function and the initialization later in this section.

In the query phase, we have a query graph $G_q$ with query weights $w_q$ and a mapping $T$ which maps nodes from $G_q$ to $G_\ell$.
For most applications, $G_q = G_\ell$ and $T$ will be the identity.
The query weights $w_q$ are the main mechanism to implement different applications with CH-Potentials.
In some cases, $w_q$ will be simple scalar values like the lower bound weights.
However, for $w_q$ this is not a requirement.
It can be any function that computes a weight for an edge.
This function can also take additional parameters from the state of the search.

The query phase then is to execute A* on $G_q$.
Before running the search, the heuristic will be initialized with the target $T(t)$.
The distance estimate for a node $x$ from $G_q$ can be computed as $h_{T(t)}(T(x))$.
For CH-Potentials to find correct distances, $w_q \geq w_\ell$ has to hold for all arcs.
\todo{Correctness Argument?}
\todo{Describe A*?}

>>>>

\section{Introduction}
\label{sec:intro}
The past decade has seen a plethora of research on route planning in large street networks~\cite{bdgmpsww-rptn-16}.
Routing a user through a road network can be formalized as the shortest path problem in weighted graphs.
Nodes represent intersections.
Roads are modeled using edges.
Edges are weighted by their traversal times.
The problem can be solved with Dijkstra's algorithm~\cite{d-ntpcg-59}.
Unfortunately, on continental sized networks, it is too slow for many applications.
Thus, speed-up techniques have been developed.
One popular example are Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
They have been used successfully in many real world applications\todo{citation?}.
A CH exploits the inherent hierarchy of road networks.
In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network during query time.
A further popular example is Multi-Level-Dijkstra~(MLD)~\todo{pre CRP zitat?} also known as CRP~\cite{dgpw-crprn-13}.
It is also used in practice~\todo{citation, bing blog?}.
MLD also also uses shortcut edges.
Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.

@@

Unfortunately, for many real world applications, this basic graph model is too simplistic.
For realistic routing, many additional features need to be considered.
This includes turn costs and restrictions, live traffic, user preferences, and traffic predictions.%, temporary driving bans for certain types of vehicles.
Some applications may have additional application-specific requirements.
Extending Dijkstra's algorithm to support these features is usually easy.
Extending hierarchical speed-up techniques is also possible.
However, the algorithm development is vastly more complex.
For every feature, dedicated research paper(s) exist that extend CH.
Supporting the combination of several features is even harder.
For example, we are not aware of any work combining all features mentioned above.
In this paper, we describe an algorithmic building block, that allows handling the combination of all above mentioned features -- and probably more.

Our approach decouples extensions from the hierarchical speed-up technique by utilizing the A* algorithm~\cite{}.
A* is a goal-directed variant of Dijkstra's algorithm.
See Figure~\ref{img:search-space} for an example of nodes traversed during an A* search.
A* uses a \emph{heuristic} to guide the search towards the goal.
A heuristic is function that maps a node $v$ onto an estimate of the distance from $v$ to the goal.
A*'s running time crucially depends on how tight this estimate is.
Further, evaluating the heuristic must be fast.
% The free flow travel time to the goal is a valid heuristic for most features.
% In many settings, it is the tightest lower bound available during preprocessing, i.e., independent of real-time data and user settings.
% However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
In this paper, we describe CH-Potentials, a heuristic that achieves both goals.
Internally, the heuristic uses a CH.
Fortunately, this is an implementation detail from the perspective of the A*.
To support a new feature, we only need to modify the A* algorithm.
The heuristic core containing the CH remains untouched.
Extending A* is vastly easier than extending a CH.
This enables us to design algorithms that support a multitude of features.
In addition, we describe query optimizations for handling of low-degree nodes, common in road networks.
These low degree optimizations are applicable to Dijkstra's algorithm and A*.

@@

\subsection{Formal Setup: Inputs, Outputs, and Phases}

In this paper, we consider different applications, with slightly different problem models.
The goal is always to quickly answer many shortest path queries.
For the purpose of describing our framework, we establish a shared notation:
Input to each query are nodes $s$ and $t$, and a graph $G$ with query weights $w_q$.
However, the precise formal inputs of the query and what exactly $w_q$ represents depends on the application.
In the simplest case, $w_q$ will be scalar edge weights.
However, this is not a requirement.
It can be any function that computes a weight for an edge.
This function can also take additional parameters from the state of the search.
For example, in the case of live-traffic, $w_q$ represents scalar edge weights.
However, values of $w_q$ might change between queries.
In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time and the query takes an additional departure time parameter.

To enable quick shortest path computations, we consider a two phase setup with an additional off-line preprocessing phase before the on-line query phase.
The input to the preprocessing phase is a graph $G_\ell$ with lower bound weights $w_\ell$ and a node mapping function $\phi$.
$w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
We require that $w_q(uv)$ is greater or equal to the shortest distance from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
The output of the preprocessing is auxiliary data that enables an efficient heuristic function $h_t(x)$.
$h_t(x)$ is the exact distance from $\phi(x)$ to $\phi(t)$ in $G_\ell$.
In the applications considered in this paper, $w_\ell$ is always the freeflow travel time.

The query phase uses this heuristic in an A* search between nodes $s$ and $t$ on $G_q$ and $w_q$.
The exact implementation of this A* search depend on the application.
Our approach only provides the heuristic $h_t$ for the A* search.
In contrast, the preprocessing phase remains the same for all applications.
